### 2024-05-03 paper/paper.qmd
various todos

acknowledgements: |
  | Thanks to Emily Merchant and John Jackson for some initial discussions that helped clarify the scope of the project and identify some key resources related to the Pioneer Fund and *Mankind Quarterly*.  Thanks to Anthony Sainez for help retrieving the *Mankind Quarterly* articles.  Thanks to Derek Devnich and James Dooley for their work attempting to secure the APA-published articles.  Thanks to Kevin Bird, John Jackson, and Manuela Fernández Pinto for comments on a draft of this paper. 
  | **Funding:** EJCL's work on this project was partially supported by UC Merced.  DJH's work on this project was not supported by any specific funding. 
  | **Author contributions:** DJH (conceptualization, data curation, formal analysis, investigation, methodology, project administration, software, validation, visualization, writing-original draft, writing-review & editing); EJCL (conceptualization, methodology, software, validation, visualization, writing-original draft, writing-review & editing)
  | **Competing interests:** The authors have no competing interests to declare. 
  | **Data and materials availability:** Web of Science search results and article fulltext are not publicly available due to intellectual property restrictions.  Preprocessed analysis data (document-term counts and article metadata) and analysis scripts are available at *[doi]*.

bibliography: [references.bib, refs-manual.bib]
# cite-method: "citeproc"
# csl: "nature-no-superscript.csl"

format: 
  pnas-pdf:
    keep_tex: true
    latex_engine: "xelatex"
    link-citations: true
    mathspec: true
    classoption:
      - twocolumn
      - twoside
    fig.env: "figure*"

@@

<!-- 
PNAS guidelines
Research Report/Direct Submission: The preferred length of these articles is 6 pages, but PNAS allows articles up to a maximum of 12 pages. A standard 6-page article is approximately 4,000 words, 50 references, and 4 medium-size graphical elements (i.e., figures and tables).

@@

*[TODO: 
- NHB reviews todo
- sort git tree
- reorganize into PNAS ms + SI
- check cite formatting
- copy updated plots? 
- abstract formatting
- figure sizes
]*

<!-- # Introduction {.unnumbered} -->

@@

We first propose a distinction between scientific racism, race science, and race science discourse.  *Scientific racism* refers to the social practice of purporting to justify racial inequality and colonialism by appealing to the epistemic authority of science.  *Race science* refers to scientific and/or pseudoscientific research, including research products (journal articles, etc.), that can be utilized for scientific racism.  *Race science discourse* refers to a broader category, any treatment of race science as a legitimate area of scientific research.  This includes methodological and empirical critiques of race science.  For example, the Flynn effect (the finding that IQ scores have increased over time) poses a serious empirical challenge to hereditarian claims of racial differences in intelligence.  But technical debates over the extent and causes of the Flynn effect likely create the impression among the general public of ordinary scientific disagreement.  In this way technical critiques of race science based on the Flynn effect could potentially legitimize race science.  So empirical research supporting the Flynn effect is usually not itself race science, but can still be race science discourse.

These distinction allow us to bracket the intentions and attitudes of the researchers involved in any particular piece of research or debate.  For the purposes of the current study, it is not important whether, for instance, psychologist Arthur Jensen is or is not correctly labelled a scientific racist [@JacksonJr.ArthurJensenEvolutionary2022].  But it is important that much of Jensen's research was useful and indeed utilized to purportedly justify racial inequality, and so unambiguously qualifies as race science.

<!-- # Scientific racism, eugenics, and *Brown* {.unnumbered} -->

@@

In light of the scientific racist origins of Pioneer and MQ, and the significant attention MQ has received in recent historiography, we hypothesized that bibliometric and text mining analyses would show race science ideas originating in MQ, and from there being disseminated to mainstream publications.  This hypothesis was not supported.  Instead, mainstream psychology journals provided a venue for a distinct form of race science, based on intelligence research and originating outside of MQ.

@@

Figure @fig-focal and this discussion focuses on three topics identified in one of the 24 fitted topic models; see figures [@fig-gamma-lg; @fig-gamma-md; @fig-gamma-sm; @fig-silge-lg; @fig-silge-md; @fig-silge-sm] for all topics from all models, and figures [@fig-grid-lg; @fig-grid-md; @fig-grid-sm] for selected topics from all models.  In the focal model, topic 07 is strongly associated with MQ:  MQ published dozens of articles in this topic each year, and no other journal ever published more than a handful.  The "Silge plot," showing the top 15 terms in the topic [@SilgeTopicModeling2017], contains a number of racial terms (`races`, `whites`, `negroes` and potentially `europe`, `africa`, `india`, and `japan`); as well as `book`, likely reflecting the fact that MQ published a number of book reviews, while the mainstream journals either did not publish book reviews or these were not available for the corpus.

Topic 22 is strongly associated with Int and PID in the same way that 07 is associated with MQ.  The Silge plot does contain `jensen`, as well as a reference to Raymond Cattell, who played a major role in the development of factor analysis and intelligence testing but also advocated for eugenics, fascism, and Nazi race science [@MehlerBeyondismRaymondCattell1997]. However, the other authors named in this topic — John Horn and Peter Bentler — do not appear to have contributed to scientific racism.  Instead this topic appears to identify "mainstream" (non-race science) intelligence research, especially factor analysis and the debate over whether intelligence is unidimensional or multidimensional.  (Three other topics only associated with mainstream intelligence journals and terms were also identified by this topic model.)  This topic indicates that the model is able to distinguish "mainstream" intelligence research from race science discourse.

The Silge plot for topic 24, by contrast, suggests a distinct race science discourse topic, with multiple racial terms (`whites`, `blacks`, `racial_differences`, `races`, `race_differences`) and the names of three prominent Pioneer-funded researchers, `jensen`, `lynn`, and `rushton`.  Independent qualitative coding of 121 papers in this topic (those with $\gamma > 0.97$ and published in journals other than PR) confirmed this interpretation, with 108 (89%) coded as race science discourse by both authors (Cohen's $\kappa = 0.86$).

@@

## Race science, behavior genetics, and psychology {.unnumbered}

@@

However, the topic model analysis suggests that, by the 1990s, behavior genetics may have distanced itself somewhat from race science, albeit without directly repudiating it. This interpretation was supported by a supplemental analysis that focused on BG specifically (see methods and materials)<!-- [@sec-bg] -->.  In addition, the 1995 BGA presidential address by Glayde Whitney --- in which he criticized the "Marxist-Lysenkoist denial of genetics" and proposed that differences in murder rates between countries and cities were caused by racial genetic differences in intelligence, empathy, aggression, and impulsivity --- was published in MQ rather than BG. Behavior genetics may have been prominent in promoting race science in the past, but appears to have been less receptive to such ideas in more recent history.  Instead, our results suggest that other subdisciplines of psychology have provided the primary venue for mainstreaming race science.

# Discussion {.unnumbered}

Using bibliometric and text-mining methods, this study finds that Pioneer-funded researchers published heavily in certain mainstream psychology journals, much more than in *Mankind Quarterly*; and that a distinct topic of race science discourse, centered on race and intelligence, can be identified in these same mainstream journals across the period 1960-2010.  These findings indicate that *Mankind Quarterly* (MQ) was less important as venue for late 20th century race science than mainstream psychology journals, especially *Intelligence* (Int) and *Personality and Individual Differences* (PID).  Indeed, until the 1990s, MQ published a very different kind of race science from the race-and-intelligence research published in Int and PID.  During this same period of time, race scientists served in prominent and influential positions in behavior genetics and psychology, including as society presidents and members of journal editorial boards.  In some cases, some race scientists simultaneously maintained active connections to both the mainstream scientific community and white supremacist organizations [@WinstonScientificRacismNorth2020; @JacksonJr.ArthurJensenEvolutionary2022; @SlobodianUnequalMindHow2023].

@@

<!-- 
\clearpage
\pagenumbering{arabic}
\renewcommand*{\thepage}{\thesection.\arabic{page}}
\appendix
\renewcommand{\thesection}{S\arabic{section}}
\renewcommand\thefigure{\thesection.\arabic{figure}}    
\setcounter{figure}{0} 
\renewcommand\thetable{\thesection.\arabic{table}}    
\setcounter{table}{0}

\clearpage

# Supplemental Materials {.unnumbered}

| 
| S1. Materials and methods
| Figures S2-S7
| Data S8, S9

\clearpage
 -->

# Materials and methods

@@

We then used the author search tool in Clarivate's Web of Science platform (<https://www.webofscience.com/wos/author/search>), retrieving publication lists for 14 researchers.  These searches were conducted between 2021-09-24 and 2021-10-05 by *[AUTHOR 1]*.  After parsing these results, we counted how many of the 14 researchers had published in each journal.  Thirteen journals had published 6 or more of the 14 Pioneer-funded researchers.  We excluded *Mankind Quarterly* (as already included) as well as *Science* and *Nature* (as too general) from further consideration in this side of the corpus.  Three journals published by the American Psychological Association (APA; *American Psychologist*, *Contemporary Psychology*, and *Journal of Educational Psychology*) had to be excluded due to confusion over who could give us permission to use the archives for a text mining project, with both APA and ProQuest asserting that we needed to get permission from the other entity.  *European Journal of Personality*, published by SAGE, also had to be excluded because our institutional access only went back to 1999.

@@

The public analysis repository, <!-- <https://github.com/dhicks/race-science> --> *[URL and DOI will go here]*, includes document metadata and documentwise counts of NLP-extracted noun phrases ("document-term matrices"), along with code to reproduce the analysis of the following sections.

@@

To fit topic models, we followed the approach proposed by Rohe and Zeng [@RoheVintageFactorAnalysis2020], which uses varimax-rotated partial principal components instead of the variational inference methods used by standard topic model packages such as `stm` [@RobertsStmPackageStructural2019].  This novel approach was implemented in the R package *[redacted for anonymous review]* <!-- `tmfast` [@HicksTmfastFitsTopic2023] -->.  <!-- We conducted a simulation study of `tmfast`, which found that it was significantly faster and only slightly less accurate at reconstructing known word-topic and topic-document distributions, compared to `stm` [@HicksTmfastFitsTopic2023].   -->

@@

Following the approach of [@HicksProductivityInterdisciplinaryImpacts2021], we did not attempt to identify a unique best fitted model for further analysis.  While the manuscript text focuses on the medium vocabulary, $k=40$ model as the "median" among the 24 fitted models, we compare and contrast findings from this model with those from the other models.

### Proof of concept

<!-- {{< include ../scripts/11-proof-of-concept.qmd >}} -->

### Topic model interpretation and quality assessment

@@

The gamma distributions and Silge plots suggested that, once $k$ was sufficiently large, the topic models were distinguishing between two types of race science articles.  We therefore examined the prevalence of topics that used racial terms, intelligence-research terms, or both, by journal and across all 24 models; see figures [@fig-grid-lg; @fig-grid-md; @fig-grid-sm]. In these visualizations, each row of facets corresponds to a value of $k$ (excluding 5 as too noisy), and each column of facets corresponds to one topic.  Topics are clustered based on whether the top 15 terms contain racial terms ("race"), terms related to intelligence research ("intelligence"), or both.  Line plots show the count of articles with $\gamma > 0.50$ for the given topic, by journal; thick lines are 5-year running averages and thin lines are raw counts.

@@

We next conducted a topic quality check of topic 24 from the medium vocabulary, $k=40$ model, which appeared to identify race science research on intelligence published in mainstream journals.  A spreadsheet of all articles that had a maximum value of $\gamma$ for this topic was extracted (excluding articles published in *Psychological Reports*), and the top 121 articles ($\gamma > 0.97$) were reviewed manually by both authors.  We coded each article as *race science discourse*, or not, using our definition of race science discourse as treating race science as a legitimate area of scientific reearch; this includes methodological critiques of race science and empirical tests that falsify race science hypotheses.

@@

### STM robustness check

@@

### *[author analysis]*

@@

\clearpage

# Extended Data

\renewcommand{\thefigure}{ED-\arabic{figure}}
\setcounter{figure}{0}

![Fitted topic models: Gamma (topic-document) distributions, large vocabulary.](img/05_lg_gamma.png){#fig-gamma-lg width="6in" height="4.8in"}

![Fitted topic models: Gamma (topic-document) distributions, medium vocabulary.](img/05_md_gamma.png){#fig-gamma-md width="6in" height="4.8in"}

![Fitted topic models: Gamma (topic-document) distributions, small vocabulary.](img/05_sm_gamma.png){#fig-gamma-sm width="6in" height="4.8in"}

![Silge plots (top 15 terms) for each topic and model, large vocabulary.](supplements/S2-lg-silge.pdf){#fig-silge-lg}

![Silge plots (top 15 terms) for each topic and model, medium vocabulary.](supplements/S3-md-silge.pdf){#fig-silge-md}

![Silge plots (top 15 terms) for each topic and model, small vocabulary.](supplements/S4-sm-silge.pdf){#fig-silge-sm}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, large vocabulary.](supplements/S5_grid_lg.pdf){#fig-grid-lg}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, medium vocabulary.](supplements/S6_grid_md.pdf){#fig-grid-md}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, small vocabulary.](supplements/S7_grid_sm.pdf){#fig-grid-sm}

>>>>

bibliography: [references.bib, refs-manual.bib]
csl: "pnas.csl"

format: 
  pnas-pdf:
    keep_tex: true
    latex_engine: "xelatex"
    link-citations: true
    mathspec: true
    classoption:
      - twocolumn
      - twoside
    fig.env: "figure*"
    number-sections: false

@@

<!-- 
PNAS guidelines: <https://www.pnas.org/author-center/submitting-your-manuscript#article-types>

Research Report/Direct Submission: The preferred length of these articles is 6 pages, but PNAS allows articles up to a maximum of 12 pages. A standard 6-page article is approximately 4,000 words, 50 references, and 4 medium-size graphical elements (i.e., figures and tables).

@@

*[TODO: 
- NHB reviews todo
- de-anonymize
- copy updated plots? 
- figure sizes
- SI xrefs
]*

# Significance

*[120 words, "undergrad-educated scientist outside their field of specialty"]*

# Introduction

@@

We first propose a distinction between scientific racism, race science, and race science discourse.  *Scientific racism* refers to the social practice of purporting to justify racial inequality and colonialism by appealing to the epistemic authority of science.  *Race science* refers to scientific and/or pseudoscientific research, including research products (journal articles, etc.), that can be utilized for scientific racism.  *Race science discourse* refers to a broader category, any treatment of race science as a legitimate area of scientific research.

We stress that race science discourse includes methodological and empirical critiques of race science.  We argue that such technical critiques — even when they are intended to undermine race science — can instead legitimize it, by creating an impression among the broader public that race science is a matter of ordinary, mainstream scientific debate.  For example, the Flynn effect (the finding that IQ scores have increased over time) poses a serious empirical challenge to hereditarian claims of racial differences in intelligence, and has been extensively debated in the mainstream scientific literature.  Most members of the general public lack the time and skills needed to assess the evidence and arguments concerning the Flynn effect.  Instead, we suggest, they are likely to understand the Flynn effect — and, by extension, race science — as an area of active, mainstream scientific research and ordinary scientific disagreement.  In this way technical critiques of race science based on the Flynn effect can legitimize race science.  So empirical research supporting the Flynn effect is usually not itself race science, but can still be race science discourse.

The distinction between scientific racism, race science, and race science discourse allows us to bracket the intentions and attitudes of the researchers involved in any particular piece of research or debate.  For the purposes of the current study, it is not important whether, for instance, psychologist Arthur Jensen is or is not correctly labelled a scientific racist [@JacksonJr.ArthurJensenEvolutionary2022].  But it is important that much of Jensen's research was useful and indeed utilized to purportedly justify racial inequality, and so unambiguously qualifies as race science.

## Scientific racism, eugenics, and *Brown* {.unnumbered}

@@

In light of the scientific racist origins of Pioneer and MQ, and the significant attention MQ has received in recent historiography, we hypothesized that bibliometric and text mining analyses would show race science discourse originating in MQ, and from there being disseminated to mainstream publications.  This hypothesis was not supported.  Instead, mainstream psychology journals provided a venue for a distinct form of race science discourse, based on intelligence research and originating outside of MQ.

@@

Figure @fig-focal focuses on three topics identified in one of the 24 fitted topic models; see figures [@fig-gamma-lg; @fig-gamma-md; @fig-gamma-sm; @fig-silge-lg; @fig-silge-md; @fig-silge-sm] for all topics from all models, and figures [@fig-grid-lg; @fig-grid-md; @fig-grid-sm] for selected topics from all models.  In the focal model, topic 07 is strongly associated with MQ:  MQ published dozens of articles in this topic each year, and no other journal ever published more than a handful.  The "Silge plot," showing the top 15 terms in the topic [@SilgeTopicModeling2017], contains a number of racial terms (`races`, `whites`, `negroes` and potentially `europe`, `africa`, `india`, and `japan`); as well as `book`, likely reflecting the fact that MQ published a number of book reviews, while the mainstream journals either did not publish book reviews or these were not available for the corpus.

Topic 22 is strongly associated with Int and PID in the same way that 07 is associated with MQ.  The Silge plot does contain `jensen`, as well as a reference to Raymond Cattell, who played a major role in the development of factor analysis and intelligence testing but also advocated for eugenics, fascism, and Nazi race science [@MehlerBeyondismRaymondCattell1997]. However, the other authors named in this topic — John Horn and Peter Bentler — do not appear to have contributed to scientific racism.  Instead this topic appears to identify "mainstream" (non-race science discourse) intelligence research, especially factor analysis and the debate over whether intelligence is unidimensional or multidimensional.  (Three other topics only associated with mainstream intelligence journals and terms were also identified by this topic model.)  This topic indicates that the model is able to distinguish "mainstream" intelligence research from race science discourse.

The Silge plot for topic 24, by contrast, suggests a distinct race science discourse topic, with multiple racial terms (`whites`, `blacks`, `racial_differences`, `races`, `race_differences`) and the names of three prominent Pioneer-funded researchers, `jensen`, `lynn`, and `rushton`.  Independent qualitative coding of 121 papers in this topic (those with $\gamma > 0.97$ and published in journals other than PR) confirmed this interpretation, with 108 (89%) coded as race science discourse by both authors (Cohen's $\kappa = 0.86$); see @sec-topic-quality.

@@

*[across all models]*

- As $k$ increases, multiple different "intelligence only" topics
- lg, md: usually 1 "race only" topic and/or 1 race-and-intelligence topic
    - "race only" always almost exclusively in MQ
    - md: race-and-intelligence predominantly in mainstream
    - lg: for k=30, race-and-intelligence predominantly MQ; no race-int for k=40, 50; then similar to md for k=60, 70
- sm: 
    - seems to have a distinct "race only" topic in PR (V04)
        - top terms include race_differences, rushton, crime: race science, but poss. either not about intelligence or int-related terms didn't make it into top terms
    - other one is almost exclusive to MQ
    - except for k=40, race-int topic, predominantly mainstream

## Race science discourse, behavior genetics, and psychology {.unnumbered}

@@

However, the topic model analysis suggests that, by the 1990s, behavior genetics may have distanced itself somewhat from race science, albeit without directly repudiating it. This interpretation was supported by a supplemental analysis that focused on BG specifically; see @sec-bg.  In addition, the 1995 BGA presidential address by Glayde Whitney --- in which he criticized the "Marxist-Lysenkoist denial of genetics" and proposed that differences in murder rates between countries and cities were caused by racial genetic differences in intelligence, empathy, aggression, and impulsivity --- was published in MQ rather than BG. Behavior genetics may have been prominent in promoting race science in the past, but appears to have been less receptive to such ideas in more recent history.  Instead, our results suggest that other subdisciplines of psychology have provided the primary venue for mainstreaming race science.

# Discussion

Using bibliometric and text-mining methods, this study finds that Pioneer-funded researchers published heavily in certain mainstream psychology journals, much more than in *Mankind Quarterly*; and that a distinct topic of race science discourse, centered on race and intelligence, can be identified in these same mainstream journals across the period 1960-2010.  These findings indicate that *Mankind Quarterly* (MQ) was less important as venue for late 20th century race science discourse than mainstream psychology journals, especially *Intelligence* (Int) and *Personality and Individual Differences* (PID).  Indeed, until the 1990s, MQ published a very different kind of race science discourse from the race-and-intelligence research published in Int and PID.  During this same period of time, race scientists served in prominent and influential positions in behavior genetics and psychology, including as society presidents and members of journal editorial boards.  In some cases, some race scientists simultaneously maintained active connections to both the mainstream scientific community and white supremacist organizations [@WinstonScientificRacismNorth2020; @JacksonJr.ArthurJensenEvolutionary2022; @SlobodianUnequalMindHow2023].

@@

# Methods and materials

@@

We then used the author search tool in Clarivate's Web of Science platform (<https://www.webofscience.com/wos/author/search>), retrieving publication lists for 14 researchers.  These searches were conducted between 2021-09-24 and 2021-10-05 by DH.  After parsing these results, we counted how many of the 14 researchers had published in each journal.  Thirteen journals had published 6 or more of the 14 Pioneer-funded researchers.  We excluded *Mankind Quarterly* (as already included) as well as *Science* and *Nature* (as too general) from further consideration in this side of the corpus.  Three journals published by the American Psychological Association (APA; *American Psychologist*, *Contemporary Psychology*, and *Journal of Educational Psychology*) had to be excluded due to confusion over who could give us permission to use the archives for a text mining project, with both APA and ProQuest asserting that we needed to get permission from the other entity.  *European Journal of Personality*, published by SAGE, also had to be excluded because our institutional access only went back to 1999.

@@

The public analysis repository, <https://github.com/dhicks/race-science> *DOI will go here]*, includes document metadata and documentwise counts of NLP-extracted noun phrases ("document-term matrices"), along with code to reproduce the analysis of the following sections.

@@

To fit topic models, we followed the approach proposed by Rohe and Zeng [@RoheVintageFactorAnalysis2020], which uses varimax-rotated partial principal components instead of the variational inference methods used by standard topic model packages such as `stm` [@RobertsStmPackageStructural2019].  This novel approach was implemented in the R package `tmfast` [@HicksTmfastFitsTopic2023].  We conducted a simulation study of `tmfast`, which found that it was significantly faster and only slightly less accurate at reconstructing known word-topic and topic-document distributions, compared to `stm` [@HicksTmfastFitsTopic2023].  We report a proof-of-concept simulation for the current study in @sec-poc.

@@

Following the approach of [@HicksProductivityInterdisciplinaryImpacts2021], we did not attempt to identify a unique best fitted model for further analysis.  While the main text of this paper focuses on the medium vocabulary, $k=40$ model as the "median" among the 24 fitted models, qualitatively similar results appear across all models with $k \geq 10$ for all vocabularies; see @sec-topic-quality and @fig-grid-lg, @fig-grid-md, @fig-grid-sm.

We did conduct two further assessments of topic 24 from the medium vocabulary, $k = 40$ model.  A manual quality assessment achieved consensus that 89% (108/121) of the top papers in this topic were race science discourse; see @sec-topic-quality.  We also replicated the medium vocabulary, $k = 40$ model using the popular `STM` package for topic modeling; see @sec-stm.  This model also gave qualitatively similar results.  Due to the time and computational resources required for these robustness checks, we determined that it would not be feasible for us to repeat them across all 855 topics ($3 \times (5 + 10 + 20 + \cdots + 70)$) and 24 topic models.

# Acknowledgments

Thanks to Emily Merchant and John Jackson for some initial discussions that helped clarify the scope of the project and identify some key resources related to the Pioneer Fund and *Mankind Quarterly*.  Thanks to Anthony Sainez for help retrieving the *Mankind Quarterly* articles.  Thanks to Derek Devnich and James Dooley for their work attempting to secure the APA-published articles.  Thanks to Kevin Bird, John Jackson, and Manuela Fernández Pinto for comments on a draft of this paper.

**Funding:** EJCL's work on this project was partially supported by UC Merced.  DJH's work on this project was not supported by any specific funding.

**Author contributions:** DJH (conceptualization, data curation, formal analysis, investigation, methodology, project administration, software, validation, visualization, writing-original draft, writing-review & editing); EJCL (conceptualization, methodology, software, validation, visualization, writing-original draft, writing-review & editing)

**Competing interests:** The authors have no competing interests to declare.

**Data and materials availability:** Web of Science search results and article fulltext are not publicly available due to intellectual property restrictions.  Preprocessed analysis data (document-term counts and article metadata) and analysis scripts are available at *[doi]*.

\clearpage
\pagenumbering{arabic}
\renewcommand*{\thepage}{S\arabic{page}}
\renewcommand{\thesection}{\emph{SI Appendix}, section \arabic{section}}
\renewcommand\thefigure{S\arabic{figure}}    
\setcounter{figure}{0} 
\renewcommand\thetable{S\arabic{table}}    
\setcounter{table}{0}

# SI Appendix

## Proof of concept {#sec-poc}

{{< include ../scripts/11-proof-of-concept.qmd >}}

## Topic model interpretation and quality assessment {#sec-topic-quality}

@@

The gamma distributions and Silge plots suggested that, once $k$ was sufficiently large, the topic models were distinguishing between two types of race science discourse articles.  We therefore examined the prevalence of topics that used racial terms, intelligence-research terms, or both, by journal and across all 24 models; see figures [@fig-grid-lg; @fig-grid-md; @fig-grid-sm]. In these visualizations, each row of facets corresponds to a value of $k$ (excluding 5 as too noisy), and each column of facets corresponds to one topic.  Topics are clustered based on whether the top 15 terms contain racial terms ("race"), terms related to intelligence research ("intelligence"), or both.  Line plots show the count of articles with $\gamma > 0.50$ for the given topic, by journal; thick lines are 5-year running averages and thin lines are raw counts.

@@

We next conducted a topic quality check of topic 24 from the medium vocabulary, $k=40$ model, which appeared to identify race science discourse on intelligence published in mainstream journals.  A spreadsheet of all articles that had a maximum value of $\gamma$ for this topic was extracted (excluding articles published in *Psychological Reports*), and the top 121 articles ($\gamma > 0.97$) were reviewed manually by both authors.  We coded each article as *race science discourse*, or not, using our definition of race science discourse as treating race science as a legitimate area of scientific research; this includes methodological critiques of race science and empirical tests that falsify race science hypotheses.

@@

## STM robustness check {#sec-stm}

@@

@@

![Fitted topic models: Gamma (topic-document) distributions, large vocabulary.](img/05_lg_gamma.png){#fig-gamma-lg width="6in" height="4.8in" fig-env="figure*"}

![Fitted topic models: Gamma (topic-document) distributions, medium vocabulary.](img/05_md_gamma.png){#fig-gamma-md width="6in" height="4.8in" fig-env="figure*"}

![Fitted topic models: Gamma (topic-document) distributions, small vocabulary.](img/05_sm_gamma.png){#fig-gamma-sm width="6in" height="4.8in" fig-env="figure*"}

![Silge plots (top 15 terms) for each topic and model, large vocabulary.](supplements/S2-lg-silge.pdf){#fig-silge-lg fig-env="figure*"}

![Silge plots (top 15 terms) for each topic and model, medium vocabulary.](supplements/S3-md-silge.pdf){#fig-silge-md fig-env="figure*"}

![Silge plots (top 15 terms) for each topic and model, small vocabulary.](supplements/S4-sm-silge.pdf){#fig-silge-sm fig-env="figure*"}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, large vocabulary.](supplements/S5_grid_lg.pdf){#fig-grid-lg fig-env="figure*"}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, medium vocabulary.](supplements/S6_grid_md.pdf){#fig-grid-md fig-env="figure*"}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, small vocabulary.](supplements/S7_grid_sm.pdf){#fig-grid-sm fig-env="figure*"}

### 2024-05-03 scripts/11-proof-of-concept.qmd
various todos

---
format:
  - html
  - pdf: 
      mathspec: true
---

@@

tidy(fitted, k = 20, matrix = 'gamma') |> 
    add_journal() |> 
    ggplot(aes(document, topic, fill = gamma)) +
    geom_raster() +
    facet_wrap(vars(journal), scales = 'free_x', 
               ncol = 7) +
    scale_y_discrete(breaks = c('V01', 'V10', 'V20')) +
    scale_x_discrete(breaks = 'none')
```

@@

```{r}
#| fig-cap: "In the proof-of-concept simulation, the topic model clearly detects the marginal topic in journal A, increase in journal B over time, and minimal presence in other journals"
tidy(fitted, k = 20, matrix = 'gamma') |> 
    filter(topic == 'V01') |> 
    add_journal() |> 
    ggplot(aes(journal, gamma)) +
    ggforce::geom_sina() +
    stat_summary(color = 'red')
```

@@

```{r}
#| fig-cap: "In the proof-of-concept simulation, the topic model does not produce a false negative pattern with the mainstream topics"
tidy(fitted, k = 20, matrix = 'gamma') |> 
    filter(topic == 'V02') |> 
    add_journal() |> 
    ggplot(aes(journal, gamma)) +
    geom_sina() +
    stat_summary(color = 'red')
```

@@

```{r}
#| fig-cap: "In the proof-of-concept simulation, without normalization the estimated topic-document probabilities (gamma) are well below the true probabilities (theta) for approx. $\\theta > 0.05$"
tidy(fitted, k = 20, matrix = 'gamma') |> 
    filter(topic == 'V01') |> 
    add_journal() |> 
    inner_join(theta_df, by = c('document' = 'doc', 'topic')) |> 
    ggplot(aes(theta, gamma, color = journal)) +
    geom_point() +
    stat_function(fun = identity, 
                  inherit.aes = FALSE, color = 'black') +
    scale_color_viridis_d(guide = 'none')
```

>>>>

---
format:
  html: default
  pdf: 
    mathspec: true
---

@@

tidy(fitted, k = 20, matrix = 'gamma') |> 
    add_journal() |> 
    ggplot(aes(document, topic, fill = gamma)) +
    geom_raster() +
    facet_wrap(vars(journal), scales = 'free_x', 
               ncol = 7) +
    scale_y_discrete(breaks = c('V01', 'V10', 'V20')) +
    scale_x_discrete(breaks = 'none') +
    theme_bw()
```

@@

```{r}
#| fig-cap: "In the proof-of-concept simulation, the topic model clearly detects the marginal topic in journal A, increase in journal B over time, and minimal presence in other journals"
tidy(fitted, k = 20, matrix = 'gamma') |> 
    filter(topic == 'V01') |> 
    add_journal() |> 
    ggplot(aes(journal, gamma)) +
    ggforce::geom_sina() +
    stat_summary(color = 'red') +
    theme_bw()
```

@@

```{r}
#| fig-cap: "In the proof-of-concept simulation, the topic model does not produce a false negative pattern with the mainstream topics"
tidy(fitted, k = 20, matrix = 'gamma') |> 
    filter(topic == 'V02') |> 
    add_journal() |> 
    ggplot(aes(journal, gamma)) +
    geom_sina() +
    stat_summary(color = 'red') +
    theme_bw()
```

@@

```{r}
#| fig-cap: "In the proof-of-concept simulation, without normalization the estimated topic-document probabilities (gamma) are well below the true probabilities (theta) for approx. $\\theta > 0.05$"
tidy(fitted, k = 20, matrix = 'gamma') |> 
    filter(topic == 'V01') |> 
    add_journal() |> 
    inner_join(theta_df, by = c('document' = 'doc', 'topic')) |> 
    ggplot(aes(theta, gamma, color = journal)) +
    geom_point() +
    stat_function(fun = identity, 
                  inherit.aes = FALSE, color = 'black') +
    scale_color_viridis_d(guide = 'none') +
    theme_bw()
```
